## Dplyr 

```{r}
setwd("~/workshops/data-manip-viz")
```

* need to load csv surveys

```{r}
surveys <- read.csv('data/surveys.csv')

```

* Packages in R are sets of additional functions 

```{r, eval = FALSE, purl = FALSE}
#install.packages("dplyr") ## install
```

* Choose a CRAN mirror

```{r, message = FALSE, purl = FALSE}
library("dplyr")          ## load
```

* `dplyr` provides tools for the most common data manipulation tasks
* faster `plyr` - written in C
* besides working with flat files, can work with external DBs
* `dplyr` functions: `select()`,`filter()`, `mutate()`, `group_by()`, and `summarize()`

* select selects columns to keep

```{r, results = 'hide', purl = FALSE}
select(surveys, plot_id, species_id, weight)
```

* choose rows with `filter`

```{r, results = 'hide', purl = FALSE}
filter(surveys, year == 1995)
```

### Pipes

* allows combining functions togther 
* uses `%>%` as a pipe - from the `magrittr` package installed as part of `dplyr`
* Takes the output of one function and send it directly to the next

```{r, purl = FALSE}
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)
```

* assign to a new object 

```{r, results = 'hide', purl = FALSE}
surveys_sml <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

surveys_sml
```

> ### Challenge {.challenge}
>
> Using pipes, subset the data to include rows before 1995. Retain columns
> `year`, `sex`, and `weight.`

```{r, results = 'hide'}
surveys %>%
  filter(year < 1995) %>%
  select(year, sex, weight)
```
### Mutate

* create new columns based on the values in existing columns
* e.g. unit conversions or find the ratio of values in two columns

```{r, results = 'hide',purl = FALSE}
#weight is in grams
surveys %>%
  mutate(weight_kg = weight / 1000)
```

* a new column of weight in kg (let;s look `head(surveys)`)

```{r, results = 'hide', purl = FALSE}
surveys %>%
  mutate(weight_kg = weight / 1000) 
```

```{r, results = 'hide', purl = FALSE}
surveys %>%
  mutate(weight_kg = weight / 1000)
```

* first few rows are full of NAs

```{r, purl = FALSE}
surveys %>%
  mutate(weight_kg = weight / 1000) %>%
  filter(!is.na(weight)) %>%
  head
```

### Split-apply-combine 

* split the data into groups, apply some analysis to each group, and then combine the results
* use `group_by`
* show image

```{r, purl = FALSE}
surveys %>%
  group_by(sex) %>%
  tally()
```

* `group_by()` is often used together with `summarize()` which collapses each group

```{r, results = 'hide', purl = FALSE}
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))
```

* You can group by multiple columns too:

```{r, results = 'hide', purl = FALSE}
surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))

```

* most species were never weighed
* let's discard rows where `mean_weight` is `NaN` (`NaN` refers to "Not a Number")

```{r, purl = FALSE}
surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE)) %>%
  filter(!is.nan(mean_weight))
```

* note: abve `dplyr` has changed our `data.frame` to a `tbl_df` more concise output
* summarize multiple variables at the same time:

```{r, purl = FALSE}
surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE),
            min_weight = min(weight, na.rm = TRUE)) %>%
  filter(!is.nan(mean_weight))
```

> ### Challenge {.challenge}
>
> How many times was each `plot_type` surveyed?

```{r}
surveys %>% 
  group_by(plot_id) %>% 
  tally
  
```

> ### Challenge {.challenge}
>
> Use `group_by()` and `summarize()` to find the mean, min, and max hindfoot
> length for each species.

```{r}

```


> ### Challenge {.challenge}
>
> What was the heaviest animal measured in each year? Return the columns `year`,
> `genus`, `species`, and `weight`.

```{r}

```


## ggplot2


*basic ploting capabilities are available in basic R, but ggplot2 adds more powerful plotting

```{r}
# plotting package
library(ggplot2)
# piping / chaining
library(magrittr)
# modern dataframe manipulations
library(dplyr)
```

* load data directly from figshare.

```{r}
surveys_raw <- read.csv("http://files.figshare.com/1919744/surveys.csv")
```

* `surveys.csv` data contains some measurements of the animals caught in plots.

* summary

```{r}
summary(surveys_raw)
```

* we need to clean the dataset
* missing `species_id` in some records we need to remove 

```{r}
surveys <- surveys_raw %>%
           filter(species_id != "")
```

* many species with low counts, let's remove the ones below 10 counts

```{r}
# count records per species
species_counts <- surveys %>%
                  group_by(species_id) %>%
                  tally

# get names of those frequent species
frequent_species <- species_counts %>%
                    filter(n >= 10) %>%
                    select(species_id)

#use frequent_species to filter surveys
surveys <- surveys %>%
           filter(species_id %in% frequent_species$species_id)
```
```{r}
summary(surveys)
```

* summary shows NA's in weight and hindfoot_length -- Let's remove rows with missing weights.

```{r}
surveys_weight_present <- surveys %>%
                      filter(!is.na(weight))
```

> ### Challenge
>
> - Do the same to remove rows without `hindfoot_length`. Save results in the new dataframe.


```{r}
surveys_length_present <- surveys %>%
                      filter(!is.na(hindfoot_length))
```

- How would you get the dataframe without missing values?

```{r}
surveys_complete <- surveys_weight_present %>%
                    filter(!is.na(hindfoot_length))
```

> We can chain filtering together using pipe operator (`%>%`) introduced earlier.

```{r}
surveys_complete <- surveys %>%
                    filter(!is.na(weight)) %>%
                    filter(!is.na(hindfoot_length))
```

> Make simple scatter plot of `hindfoot_length` (in millimeters) as a function of
> `weight` (in grams), using basic R plotting capabilities.

```{r}
plot(x=surveys_complete$weight, y=surveys_complete$hindfoot_length)
```

## Plotting with ggplot2

* make the same plot using `ggplot2` package.
* `ggplot2` is a plotting package that makes it easier to create complex plots
* implements a grammar of graphics, so we can build up plots
* ggplot requires:
  1. bind plot to a specific data frame

```{r, eval=FALSE}
ggplot(surveys_complete)
```

  2. define aestetics (`aes`), that maps variables in the data to axes on the plot or to plotting size, shape color, etc.,

```{r, eval=FALSE}
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length))
```

  3. add `geoms` -- graphical representation of the data in the plot (points, lines, bars). To add a geom to the plot use `+` operator:

```{r}
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point()
```

## Modifying plots

* adding transparency (alpha)

```{r}
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha=0.1)
```

* adding colors

```{r}
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha=0.1, color="blue")
```

## Boxplot

* Visualising the distribution of weight within each species.

```{r}
ggplot(surveys_weight_present, aes(factor(species_id), weight)) +
                   geom_boxplot()
```

* adding points to boxplot and a jitter

```{r}
ggplot(surveys_weight_present, aes(factor(species_id), weight)) +
                   geom_jitter(alpha=0.3, color="tomato") +
                   geom_boxplot(alpha=0)
```

> ### Challenge
>
> Create boxplot for `hindfoot_length`.

```{r}
ggplot(surveys_complete, aes(factor(species_id), hindfoot_length)) +
  geom_boxplot()
```

## Time series

* counts per year for each species (group data first and count records within each group)

```{r}
yearly_counts <- surveys %>%
                 group_by(year, species_id) %>%
                 tally
```

* Timelapse data can be visualised as a line plot with years on x axis and counts on y.

```{r}
ggplot(yearly_counts, aes(x=year, y=n)) +
                  geom_line()
```

* does not work -- plots data for all the species together
* need to ggplot to split graphed data by `species_id`

```{r}
ggplot(yearly_counts, aes(x=year, y=n, group=species_id)) +
  geom_line()
```

* add colors to distinguish species

```{r}
ggplot(yearly_counts, aes(x=year, y=n, group=species_id, color=species_id)) +
  geom_line()
```

## Faceting

* allows to split one plot into mutliple plots based on some factor

```{r}
ggplot(yearly_counts, aes(x=year, y=n, color=species_id)) +
  geom_line() + facet_wrap(~species_id)
```

* we want to two lines each plot by sex 
* make counts in dataframe grouped by sex.

> ### Challenges:
>
> - filter the dataframe so that we only keep records with sex "F" or "M"s
>

```{r}
sex_values = c("F", "M")
surveys <- surveys %>%
           filter(sex %in% sex_values)
```

> - group by year, species_id, sex

```{r}
yearly_sex_counts <- surveys %>%
                     group_by(year, species_id, sex) %>%
                     tally
```

> - make the faceted plot spliting further by sex (within single plot)

```{r}
ggplot(yearly_sex_counts, aes(x=year, y=n, color=species_id, group=sex)) +
  geom_line() + facet_wrap(~ species_id)
```

> We can improve the plot by coloring by sex instead of species (species are
> already in separate plots, so we don't need to distinguish them better)

```{r}
ggplot(yearly_sex_counts, aes(x=year, y=n, color=sex, group=sex)) +
  geom_line() + facet_wrap(~ species_id)
```

