## Dplyr 

```{r}
setwd("~/workshops/data-manip-viz")
```

* need to load csv surveys

```{r}
surveys <- read.csv('data/surveys.csv')

```

* Packages in R are sets of additional functions 

```{r, eval = FALSE, purl = FALSE}
#install.packages("dplyr") ## install
```

* Choose a CRAN mirror

```{r, message = FALSE, purl = FALSE}
library("dplyr")          ## load
```

* `dplyr` provides tools for the most common data manipulation tasks
* faster `plyr` - written in C
* besides working with flat files, can work with external DBs
* `dplyr` functions: `select()`,`filter()`, `mutate()`, `group_by()`, and `summarize()`

* select selects columns to keep

```{r, results = 'hide', purl = FALSE}
select(surveys, plot_id, species_id, weight)
```

* choose rows with `filter`

```{r, results = 'hide', purl = FALSE}
filter(surveys, year == 1995)
```

### Pipes

* allows combining functions togther 
* uses `%>%` as a pipe - from the `magrittr` package installed as part of `dplyr`
* Takes the output of one function and send it directly to the next

```{r, purl = FALSE}
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)
```

* assign to a new object 

```{r, purl = FALSE}
surveys_sml <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

surveys_sml
```

> ### Challenge {.challenge}
>
> Using pipes, subset the data to include rows before 1995. Retain columns
> `year`, `sex`, and `weight.`

```{r}
surveys %>%
  filter(year < 1995) %>%
  select(year, sex, weight)
```
### Mutate

* create new columns based on the values in existing columns
* unit conversions or find the ratio of values in two columns

```{r, purl = FALSE}
surveys %>%
  mutate(weight_kg = weight / 1000)
```

* a new column of weight in kg (let;s look `head(surveys)`)

```{r, purl = FALSE}
surveys %>%
  mutate(weight_kg = weight / 1000) 
```

```{r, purl = FALSE}
surveys %>%
  mutate(weight_kg = weight / 1000) %>%
  head
```

* first few rows are full of NAs

```{r, purl = FALSE}
surveys %>%
  mutate(weight_kg = weight / 1000) %>%
  filter(!is.na(weight)) %>%
  head
```

### Split-apply-combine 

*split the data into groups, apply some analysis to each group, and then combine the results
* use `group_by`
* show image

```{r, purl = FALSE}
surveys %>%
  group_by(sex) %>%
  tally()
```

* `group_by()` is often used together with `summarize()` which collapses each group

```{r, purl = FALSE}
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))
```

* You can group by multiple columns too:

```{r, purl = FALSE}
surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))

```

* most species were never weighed
* let's discard rows where `mean_weight` is `NaN` (`NaN` refers to "Not a Number")

```{r, purl = FALSE}
surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE)) %>%
  filter(!is.nan(mean_weight))
```

* note: abve `dplyr` has changed our `data.frame` to a `tbl_df` more concise output
* summarize multiple variables at the same time:

```{r, purl = FALSE}
surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE),
            min_weight = min(weight, na.rm = TRUE)) %>%
  filter(!is.nan(mean_weight))
```

> ### Challenge {.challenge}
>
> How many times was each `plot_type` surveyed?

```{r}
surveys %>% 
  group_by(plot_id) %>% 
  tally
  
```

> ### Challenge {.challenge}
>
> Use `group_by()` and `summarize()` to find the mean, min, and max hindfoot
> length for each species.


> ### Challenge {.challenge}
>
> What was the heaviest animal measured in each year? Return the columns `year`,
> `genus`, `species`, and `weight`.

## GGPLOT2


```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(fig.keep='last')
```

```{r setup, echo=FALSE, purl=FALSE}
source("setup.R")
```

There are basic ploting capabilities in basic R, but ggplot2 adds more powerful plotting capabilities.


